\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-08 doc/incertae_sedis.tex
citing some of the early MRP stuff, and introducing the matrix rep that might help elsewhere.

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\linespread{1.5}
\usepackage[authoryear]{natbib}

@@

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.
% \paragraph{Definition of rooted splits}
As is typically done, we define rooted splits for each node by noting that
removing the edge connecting the node to its parent would produce two trees.
The leaf label sets of these trees
the partition the full leaf set $\mathcal{L}$ into two groups.
The include set $\mathcal{I}(n)$ which
does not contain the root, and the exclude set $\mathcal{E}(n)$ which does contain the root.
Such a split may be written
\[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
where the $\bullet$ indicates the root. 
The include set is the set of labels for leaves that are
descendants of the node.
The exclude set is the set of
labels for leaves that are not descendants of that node. 
If no taxa are \emph{incertae sedis},
then the exclude set for a node is just the total tip set minus the include set
for the node: $\mathcal{E}(n)  =\mathcal{L}-\mathcal{I}(n)$.
%For a node $n$ on the tip-ward side of an edge $e$, we
%may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$.
For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

@@

@@

However, there are cases where thus procedure does not result in one name per
node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
``inter-digitated'', or  combined into a single clade in such a way that their
members cannot be separated.  When the members of the two clades share the same
MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
apply to the same node.  However, in this case, neither name is ancestral to
the other, and the previous solution does not apply.
In such cases, we treat all applicable names as synonyms for the node.
\mthcomment{Need to verify that the namer records this situation as synonyms!}

>>>>

\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=3cm}
\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}
\linespread{1.0}
\usepackage{multirow}

\usepackage[authoryear]{natbib}

@@

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.
% \paragraph{Definition of rooted splits}
As is typically done, we define rooted splits for each node by noting that
removing the edge connecting the node to its parent would produce two trees.
The leaf label sets of these trees
the partition the full leaf set $\mathcal{L}$ into two groups.
The include set $\mathcal{I}(n)$ which
does not contain the root, and the exclude set $\mathcal{E}(n)$ which does contain the root.
Such a split may be written
\[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
where the $\bullet$ indicates the root. 
The include set is the set of labels for leaves that are
descendants of the node.
The exclude set is the set of
labels for leaves that are not descendants of that node.
The ``include set'' and ``exclude set'' of a ``rooted split'' used here are equivalent
    to the ``inside set'' and ``outside set'' of a ``$n$-taxon statement''
    \citep[\emph{sensu}][]{Wilkinson1994}.

If no taxa are \emph{incertae sedis},
then the exclude set for a node is just the total tip set minus the include set
for the node: $\mathcal{E}(n)  =\mathcal{L}-\mathcal{I}(n)$.
%For a node $n$ on the tip-ward side of an edge $e$, we
%may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$.
For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

@@

\paragraph{Equivalence rooted split definitions with character encodings}
Representing a split in a tree by a character is the basis of one of the oldest
    supertree methods: matrix representation with parsimony \citep[MRP, see][]{Baum1992,Ragan1992}.
In the rooted form of such an approach, character state 1 is used for tips that are
    descendants of a internal node in an input tree and 0 is used for other tips.
The partial splits caused by \emph{incertae sedis} annotations naturally map to
    an encoding of missing data (typically denoted with the `?' symbol in a matrix).
Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented as the matrix shown in 
    Figure \ref{fig:Handling-incertae-sedis-mrp}.
In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters
   in the row for taxa $C1$, $C2$, and $C3$ would be replaced with 0's.

\begin{figure}
\begin{center}
\begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
\hline
& \multicolumn{4}{|c|}{Character/clade label} \\
leaves & C & AB & B & A \\
\hline
$A1$, $A2$ & 0 & 1 & 0 & 1 \\
$B1$, $B2$ & 0 & 1 & 1 & 0 \\
$C1$, $C2$, $C3$ & 1 & ? & ? & ?  \\
$D$  & 0 & 0 & 0 & 0 \\
\hline
\end{tabular}
\end{center}
\caption{\label{fig:Handling-incertae-sedis-mrp} The matrix encoding 
of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
Only the unique character encodings are show, thus the leftmost column
lists all of the tip taxa that would be assigned a particular character encoding.
}

@@

However, there are cases where thus procedure does not result in one name per
node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
``inter-digitated'', or  combined into a single clade in such a way that their
members cannot be separated.  When the members of the two clades share the same
MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
apply to the same node.  However, in this case, neither name is ancestral to
the other, and the previous solution does not apply.
In such cases, we treat all applicable names as synonyms for the node.
\mthcomment{Need to verify that the namer records this situation as synonyms, as it should!}
